---
title: 'ECO 395M: StatLearning Exercise 1'
author: "Joseph Williams, Aahil Navroz, Suqian Qi"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    latex_engine: pdflatex
---

```{r, message=FALSE, echo=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(class)
library(rsample)
```

## 1) Data visualization: flights at ABIA

For this question we wanted to help fliers (say fliers in 2009) build intuition or 'rules of thumb' they can use when choosing between airline companies. To begin we want the data to correspond to recognizable names, so we mapped UniqueCarrier to airline brands or their parent brands.  Here's a breakdown of which major carriers are running the most flights out of ABIA.


```{r, message=FALSE, echo=FALSE, warning=FALSE}
abia_data = read.csv('./data/ABIA.csv', header = TRUE)

#Mapping to tie airlines to carrier codes
carrier_lookup = data.frame(
  UniqueCarrier = sort(unique(abia_data$UniqueCarrier)),
  Airline = c("Pinnacle", "American", "JetBlue", "North-Western", "Delta", "Alaska", "Frontier", "American", "Celeste", "American", "SkyWest", "United", "SilkAir", "Southwest", "JSX", "United")
)

#Merge code to airline name
abia_data = merge(abia_data, carrier_lookup, by = "UniqueCarrier")

#Lets check using airline == Southwest
data_check = abia_data %>% 
  select(UniqueCarrier, Airline) %>% 
  filter(Airline == "Southwest")

#Lets see what the most popular major airlines are
main_airlines = c("United", "Southwest", "JetBlue", "Delta", "American", "Frontier")

flights_by_airline = abia_data %>% 
  filter(Airline %in% main_airlines) %>%
  group_by(Airline) %>%
  summarize(TotalFlights = n())


# Use scale_fill_manual() with the vector of colors
ggplot(flights_by_airline, aes(x = Airline, y = TotalFlights, fill = Airline)) +
  geom_col() +
  labs(title = "Total Flights by Major Airline",
       x = "Airline",
       y = "Flights") +
  theme_minimal()

```


Looks like American and Southwest are king... but can they handle the volume!?  Lets look at arrival delays for each company. Given these are arrival delays for flights coming into and out of Austin, overall it will be a fine measure for the timeliness of the airline.

```{r, message=FALSE, echo=FALSE, warning=FALSE}

# Converting DepTime to hour of the day
abia_data$DepHour = floor(abia_data$DepTime / 100)

# Calculate average departure delays by airline and hour
avg_delays = abia_data %>%
  filter(Airline %in% main_airlines) %>%
  group_by(Airline, DepHour) %>%
  summarize(AvgArrDelay = mean(ArrDelay, na.rm = TRUE))

# Plotting average Arrival delay by airline across different hours of the day 
ggplot(avg_delays, aes(x = DepHour, y = AvgArrDelay, color = Airline)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Airline, scales = "fixed", ncol = 3) + 
  theme_minimal() +
  labs(title = "Average Arrival Delays by Airline and Hour",
       x = "Hour of Day",
       y = "Average Delay (Min)") +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) + 
  theme(legend.position = "none")

```

Okay, we're seeing some detail here.  Seems like most companies experience their delays before the hour of 6am.  Since we're looking for rules of thumb.  Lets classify into 'Red Eye' 'Early Morning', '9-5' and 'Evening-Night', and see if we can quantify delay times over periods, rather than specific times.  Some of these averages seem too high, too, lets remove observations where ArrDelay is more than 4 hours, since that usually results in a changed flight for me.  Lets also subtract WeatherDelay, SecurityDelay, and NASDelay from ArrivalDelay since those aren't related to the airline.


```{r, message=FALSE, echo=FALSE, warning=FALSE}

abia_data$DepHour = as.integer(abia_data$DepHour)
#missing_values = sum(is.na(abia_data$DepHour))
#print(missing_values)

abia_data = abia_data[complete.cases(abia_data$DepHour), ]
abia_data = abia_data %>% filter(DepDelay <= 240)

classify_hour = function(hour) {
  ifelse((hour >= 22 & hour < 24) | (hour >= 0 & hour < 2), "Red-Eye",
         ifelse(hour >= 2 & hour < 9, "Morning",
                ifelse(hour >= 9 & hour < 17, "9-5", "Evening-Night")))
}

abia_data = abia_data %>% 
  mutate(Period = classify_hour(DepHour))


avg_delays_period = abia_data %>%
  filter(Airline %in% main_airlines) %>%
  mutate(WeatherDelay = coalesce(WeatherDelay, 0)) %>%
  mutate(SecurityDelay = coalesce(SecurityDelay, 0)) %>%
  mutate(NASDelay = coalesce(NASDelay, 0)) %>%
  group_by(Airline, Period) %>%
  summarize(AvgArrDelay = mean(ArrDelay - WeatherDelay - SecurityDelay - NASDelay, na.rm = TRUE))



# Define the desired order of the levels for the "Period" variable
desired_order = c("Morning", "9-5", "Evening-Night", "Red-Eye")

# Reorder the levels of the "Period" variable
avg_delays_period = avg_delays_period %>%
  mutate(Period = factor(Period, levels = desired_order))

ggplot(avg_delays_period, aes(x = Period, y = AvgArrDelay, fill = Airline)) +
  geom_col() +
  facet_wrap(~Airline, scales = "fixed", ncol = 3) + 
  theme_minimal() +
  labs(title = "Average Arrival Delays by Airline and Period",
       x = "Period",
       y = "Average Delay (Min)") + 
theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

```

Okay now this is more useful!  Let's see, if cost difference is minimal and its important that my flight goes smoothly, I am brand-indifferent for Morning and 9-5 flights, prefer Frontier for night flights, and will only fly United for red-eyes.  You're welcome, 2009 travelers!

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# See what is the main reason for delay for main airlines. Delete the cancelled airlines and miss data
temp = abia_data %>%
  filter(Cancelled < 1)

temp = temp %>%
  filter(!is.na(CarrierDelay) & !is.na(WeatherDelay) & !is.na(NASDelay) & !is.na(SecurityDelay) & !is.na(LateAircraftDelay))

temp = temp %>% 
  filter(Airline %in% main_airlines)

# summarize the frequency for each reason 
delay_reason_fre = temp %>%
  group_by(Airline) %>%
  mutate(across(c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay), ~sum(. != 0)))

delay_reason_fre = delay_reason_fre %>%
  select(Airline, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay) %>%
  distinct() 

delay_reason_fre_long = tidyr::pivot_longer(delay_reason_fre, cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay), names_to = "Variable")

ggplot(delay_reason_fre_long, aes(x = Variable, y = value, fill = Variable)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  labs(title = "Frequency of Arrival Delays by Reason and Airline",
       x = "Reason",
       y = "Count") + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 60, hjust = 1)) +
  facet_wrap(~ Airline, scales = "free")
```

Through the plot we can see for American, Delta, Frontier and JetBlue, the most common reason for delay is NASDelay. For Southwest, the most common reason is LateAircraftDelay and for United it's CarrierDelay.


```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplot(delay_reason_fre_long, aes(x = Variable, y = value, fill = Variable)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  labs(title = "Frequency of Arrival Delays by Reason",
       x = "Reason",
       y = "Count") + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 60, hjust = 1))
```

However, considering the far ahead flight amount of Southwest Airlines, the overall most common reason for delay is LateAircraftDelay. Now, let's see the proportion of delay time for each reason whether corresponding to their frequency.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Calculate the percentage of each delay reason to the total arrival delay
delay_summarized = temp %>%
  group_by(Airline) %>%
  summarise(across(c('CarrierDelay', 'WeatherDelay', 'NASDelay', 'SecurityDelay', 'LateAircraftDelay'), sum))

delay_summarized = delay_summarized %>%
  mutate(total = rowSums(select(., CarrierDelay:LateAircraftDelay))) %>%
  mutate_at(vars(CarrierDelay:LateAircraftDelay), ~./total) %>%
  select(-total)

delay_summarized_long = tidyr::pivot_longer(delay_summarized, cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay), names_to = "Variable")

# Show the percentage in bar plots
ggplot(delay_summarized_long, aes(x = Variable, y = value, fill = Variable)) +
  geom_bar(stat = "identity") +
  labs(title = "Proportion of Arrival Delay time by Reason and Airline",
       x = "Reason",
       y = "Proportion") +
  facet_wrap(~ Airline, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 60, hjust = 1)) 
```

We can see for American, Delta and JetBlue, the LateAircraftDelay takes the most part of delay instead of NASDelay while the others remains the same. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
delay_summarized_all = temp %>%
  summarise(across(c('CarrierDelay', 'WeatherDelay', 'NASDelay', 'SecurityDelay', 'LateAircraftDelay'), sum))

delay_summarized_all = delay_summarized_all %>%
  mutate(total = rowSums(select(., CarrierDelay:LateAircraftDelay))) %>%
  mutate_at(vars(CarrierDelay:LateAircraftDelay), ~./total) %>%
  select(-total)

delay_summarized_all_long = tidyr::pivot_longer(delay_summarized_all, cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay), names_to = "Variable")
# Show the percentage in bar plots
ggplot(delay_summarized_all_long, aes(x = Variable, y = value, fill = Variable)) +
  geom_bar(stat = "identity") +
  labs(title = "Proportion of Arrival Delay time by Reason",
       x = "Reason",
       y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 60, hjust = 1)) 
```

With no surprise, LateAircraftDealy is the most common reason for all flights. However, all these conclusion may not be valid, since about 85% of the data are missing for delay reason. All our research are based on the rest 15% of the overall data.

### 2) Wrangling the Olympics

A) What is the 95th percentile of heights for female competitors across all Athletics events (i.e., track and field)?  

```{r, message=FALSE, echo=FALSE, warning=FALSE}
olympics_data = read.csv("./data/olympics_top20.csv")

#A

# Filter for female competitors in Athletics events
unique_female_athletes = olympics_data %>%
  filter(sex == 'F', grepl('Athletics', event)) %>%
  distinct(id, .keep_all = TRUE)

# Calculate the 95th percentile of heights
percentile_95_height = quantile(unique_female_athletes$height, probs = 0.95, na.rm = TRUE)

print(percentile_95_height)

```

The 95th percentile of heights for female competitors across all Athletics events is 183.


B) Which single women's `event` had the greatest variability in competitor's heights across the entire history of the Olympics, as measured by the standard deviation?  

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#B
females = filter(olympics_data, sex == 'F')

# Calculate the standard deviation of heights for each event
std_devs = females %>%
  group_by(event) %>%
  summarize(HeightStdDev = sd(height, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit()  

# Find the event with the greatest standard deviation in height
max_std_dev = max(std_devs$HeightStdDev, na.rm = TRUE)
event_with_greatest_variability = filter(std_devs, HeightStdDev == max_std_dev)

print(event_with_greatest_variability$event)
print(event_with_greatest_variability$HeightStdDev)
```

Rowing Women's Coxed Fours had the greatest variability in competitor's heights across the entire history of the Olympics and its corresponding standard deviation is 10.86549.


C) How has the average age of Olympic swimmers changed over time? Does the trend look different for male swimmers relative to female swimmers?

```{r, message=FALSE, echo=FALSE, warning=FALSE}
swimmers = filter(olympics_data, sport == 'Swimming')

# Calculate the average age of swimmers over time for each gender
avg_age_over_time = swimmers %>%
  group_by(year, sex) %>%
  summarize(AverageAge = mean(age, na.rm = TRUE)) %>%
  ungroup()

# Plot the data with separate lines for male and female swimmers
ggplot(avg_age_over_time, aes(x = year, y = AverageAge, color = sex)) +
  geom_line(size = 1.2) +
  labs(title = 'Average Age of Olympic Swimmers Over Time by Gender',
       x = 'Year',
       y = 'Average Age') +
  theme_minimal() 
```

Average age by gender has increased since 1920s for both type's of athletes but began to increase more rapidly for women leading up to year 2000. Suggesting women are competing at the highest level of sport later in their lives, on average.  This likely a result of increased popularity of woman's sports, effects of title IX.

### 3) K-nearest neighbors: cars  

For each trim, make a plot of RMSE versus K, so that we can see where it bottoms out.  Then for the optimal value of K, show a plot of the fitted model, i.e. predictions vs. x.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
sclass = read.csv("./data/sclass.csv")

set.seed(17)

# Filter for column trim, mileage and price
sclass = sclass[, c("trim", "mileage", "price")]

# Filter for the two trim levels
sclass_350 = sclass %>% filter(trim == '350')
sclass_65_AMG = sclass %>% filter(trim == '65 AMG')

# Split the data for the 350 trim level
split_350 = initial_split(sclass_350, prop = 0.80) 
train_350 = training(split_350)
test_350 = testing(split_350)

# Split the data for the 65 AMG trim level
split_65_AMG = initial_split(sclass_65_AMG, prop = 0.80) 
train_65_AMG = training(split_65_AMG)
test_65_AMG = testing(split_65_AMG)


# Do the KNN to calculate the RMSE for each K
run_knn = function(train_data, test_data) {
  k_values = seq(2, 100, by = 2) # Define the range of k values
  rmse_values = numeric(length(k_values)) # To store RMSE for each k
  
  for (i in seq_along(k_values)) {
    k = k_values[i]
    
    # Train the KNN model
    knn_model = knn(train = train_data[, "mileage", drop = FALSE], 
                     test = test_data[, "mileage", drop = FALSE], 
                     cl = train_data$price, 
                     k = k)
    
    # Convert knn_model to numeric
    knn_model = as.numeric(knn_model)
    
    # Calculate RMSE and store it
    rmse_values[i] = sqrt(mean((knn_model - test_data$price)^2))
  }
  
  return(data.frame(k = k_values, RMSE = rmse_values))
}


rmse_results_350 = run_knn(train_350, test_350)


rmse_results_65_AMG = run_knn(train_65_AMG, test_65_AMG)

# Plotting for 350 trim
ggplot(rmse_results_350, aes(x = k, y = RMSE)) +
  geom_line() + geom_point() +
  ggtitle("RMSE vs K for 350 Trim") +
  xlab("K") + ylab("RMSE")
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Plotting for 65 AMG trim
ggplot(rmse_results_65_AMG, aes(x = k, y = RMSE)) +
  geom_line() + geom_point() +
  ggtitle("RMSE vs K for 65 AMG Trim") +
  xlab("K") + ylab("RMSE")
```

We have shown the RMSE for different K values vary from 2 to 60 of both trims. For trim 350, the best K value is 14 and for trim 65 AMG, the best K value is 16. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}

# Identify the optimal K for 350 trim
optimal_k_350 = rmse_results_350[which.min(rmse_results_350$RMSE), ]$k

# Identify the optimal K for 65 AMG trim
optimal_k_65_AMG = rmse_results_65_AMG[which.min(rmse_results_65_AMG$RMSE), ]$k

# Function to plot predictions vs mileage for the optimal K
plot_predictions_vs_mileage = function(train_data, test_data, optimal_k, trim_name) {
  # Train the KNN model
  knn_model = knn(train = train_data[, "mileage", drop = FALSE], 
                   test = test_data[, "mileage", drop = FALSE], 
                   cl = train_data$price, 
                   k = optimal_k)
  
  # Convert knn_model to numeric
  knn_model = as.numeric(as.character(knn_model))
  
  # Create a data frame for plotting
  plot_data = data.frame(Mileage = test_data$mileage, Actual_Price = test_data$price, Predicted_Price = knn_model)
  
  ggplot(plot_data, aes(x = Mileage)) +
    geom_point(aes(y = Actual_Price), color = "blue", alpha = 0.5, size = 3.5) +
    geom_line(aes(y = Predicted_Price), color = "red", size = 1.3) +
    ggtitle(paste("Predictions vs Mileage for", trim_name, "with K =", optimal_k)) +
    xlab("Mileage") + ylab("Price") +
    theme_minimal()
}


# Plot for 350 trim
plot_predictions_vs_mileage(train_350, test_350, optimal_k_350, "350 Trim")

```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Plot for 65 AMG trim
plot_predictions_vs_mileage(train_65_AMG, test_65_AMG, optimal_k_65_AMG, "65 AMG Trim")
```

